Bayesian Analysis of Gas-free Galaxies in Simulations I: Reviving Stochasticity

### 2023-02-21 papers/paper-1/diagrams/hm_coresersic.tex
Small updates in preparation for SPZ meeting

    % define layer 2
    \node[latent, left=of S] (Sb) {$\tilde{\Sigma}_\mathrm{b}$};
    \node[latent, below=of Sb] (Re) {$R_\mathrm{e}$};
    \node[latent, below=of Re] (rb) {$r_\mathrm{b}$};
    \node[latent, above=of Sb] (g) {$\gamma$};
    \node[latent, above=of g] (n) {$n$};

    % define layer 3
    \factor[left=of rb] {rb_f} {below:$\Gamma$} {} {};
    \node[hyper, left=of rb_f, yshift=-1.2em] (rba) {$r_{\mathrm{b},\alpha}$};
    \node[hyper, left=of rb_f, yshift=1.2em] (rbb) {$r_{\mathrm{b},\beta}$};

    \factor[left=of Re] {Re_f} {below:$\Gamma$} {} {};
    \node[hyper, left=of Re_f, yshift=-1.2em] (Rea) {$R_{\mathrm{e},\alpha}$};
    \node[hyper, left=of Re_f, yshift=1.2em] (Reb) {$R_{\mathrm{e},\beta}$};

    \factor[left=of Sb] {Sb_f} {below:$\Gamma$} {} {};
    \node[hyper, left=of Sb_f, yshift=-1.2em] (Sba) {$\tilde{\Sigma}_{\mathrm{b},\alpha}$};
    \node[hyper, left=of Sb_f, yshift=1.2em] (Sbb) {$\tilde{\Sigma}_{\mathrm{b},\beta}$};

    \factor[left=of g] {g_f} {below:$\Gamma$} {} {};
    \node[hyper, left=of g_f, yshift=-1.2em] (ga) {$\gamma_\alpha$};
    \node[hyper, left=of g_f, yshift=1.2em] (gb) {$\gamma_\beta$};

    \factor[left=of n] {n_f} {below:$\Gamma$} {} {};
    \node[hyper, left=of n_f, yshift=-1.2em] (na) {$n_\alpha$};
    \node[hyper, left=of n_f, yshift=1.2em] (nb) {$n_\beta$};

@@

    \edge {rb, Re, Sb, g, n, R} {S};

    \factoredge {rba, rbb} {rb_f} {rb};
    \factoredge {Rea, Reb} {Re_f} {Re};
    \factoredge {Sba, Sbb} {Sb_f} {Sb};
    \factoredge {ga, gb} {g_f} {g};
    \factoredge {na, nb} {n_f} {n};

    % plates
    \plate {rS} {(R)(S)(sigma)(Shat)} {$\forall R,\sigma_\Sigma^2$};
    \plate {pars} {(R)(S)(sigma)(rb)(Re)(Sb)(g)(n)} {$\forall$ child};

>>>>

    % define layer 2
    \node[latent, left=of S] (Sb) {$\tilde{\Sigma}_\mathrm{b}$};
    \node[latent, below=of Sb] (Re) {$R_\mathrm{e}$};
    \node[latent, below=of Re] (rb) {$r_\mathrm{b}$};
    \node[latent, above=of Sb] (g) {$\gamma$};
    \node[latent, above=of g] (n) {$n$};
    \node[latent, above=of n] (a) {$a$};

    % define layer 3
    \factor[left=of rb] {rb_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of rb_f, yshift=-1.2em] (rbs) {$r_{\mathrm{b},\sigma}$};
    \node[hyper, left=of rb_f, yshift=1.2em] (rbm) {$r_{\mathrm{b},\mu}$};

    \factor[left=of Re] {Re_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of Re_f, yshift=-1.2em] (Res) {$R_{\mathrm{e},\sigma}$};
    \node[hyper, left=of Re_f, yshift=1.2em] (Rem) {$R_{\mathrm{e},\mu}$};

    \factor[left=of Sb] {Sb_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of Sb_f, yshift=-1.2em] (Sbs) {$\tilde{\Sigma}_{\mathrm{b},\sigma}$};
    \node[hyper, left=of Sb_f, yshift=1.2em] (Sbm) {$\tilde{\Sigma}_{\mathrm{b},\mu}$};

    \factor[left=of g] {g_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of g_f, yshift=-1.2em] (gs) {$\gamma_\sigma$};
    \node[hyper, left=of g_f, yshift=1.2em] (gm) {$\gamma_\mu$};

    \factor[left=of n] {n_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of n_f, yshift=-1.2em] (ns) {$n_\sigma$};
    \node[hyper, left=of n_f, yshift=1.2em] (nm) {$n_\mu$};

    \factor[left=of a] {a_f} {below:$\mathcal{N}$} {} {};
    \node[hyper, left=of a_f, yshift=-1.2em] (as) {$a_\sigma$};
    \node[hyper, left=of a_f, yshift=1.2em] (am) {$a_\mu$};

@@

    \edge {rb, Re, Sb, g, n, a, R} {S};

    \factoredge {rbs, rbm} {rb_f} {rb};
    \factoredge {Res, Rem} {Re_f} {Re};
    \factoredge {Sbs, Sbm} {Sb_f} {Sb};
    \factoredge {gs, gm} {g_f} {g};
    \factoredge {ns, nm} {n_f} {n};
    \factoredge {as, am} {a_f} {a};

    % plates
    \plate {rS} {(R)(S)(sigma)(Shat)} {$\forall R,\sigma_\Sigma^2$};
    \plate {pars} {(R)(S)(sigma)(rb)(Re)(Sb)(g)(n)(a)} {$\forall$ child};

### 2023-02-21 papers/paper-1/sections/results.tex
Small updates in preparation for SPZ meeting

\subsection{Projected Mass Density Profiles}
We use hierarchical modelling to recover the population parameter distributions of the core-S\'ersic model \autoref{eq:coresersic} for each simulation family. 
We enforce a sharp transition from the core to outer galaxy regions by setting $\alpha$ in \autoref{eq:coresersic} to 10.0. 
A probabilistic graphical model detailing the connection between hyperparameters, latent parameters, and observables is given in \autoref{fig:graphicalmodel}.

\begin{table}
    \centering
    \caption{The parameters for each hyperparameter distribution for the core-S\'ersic model, where each distribution is a gamma distribution.}
    \label{tab:cshypers}
    \begin{tabular}{ccc}
        \hline
        Hyperparameter & $\alpha$ & $\beta$ \\
        \hline
        $r_{\mathrm{b}, \alpha}$ & 3.0 & 2.0 \\
        $r_{\mathrm{b}, \beta}$ & 10.0 & 8.0 \\
        $R_{\mathrm{e}, \alpha}$ & 30.0 & 2.0 \\
        $R_{\mathrm{e}, \beta}$ & 40.0 & 2.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \alpha}$ & 20.0 & 4.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \beta}$ & 20.0 & 2.0 \\
        $\gamma_\alpha$ & 1.0 & 2.0 \\
        $\gamma_\beta$ & 4.0 & 2.0 \\
        $n_\alpha$ & 16.0 & 2.0 \\
        $n_\beta$ & 4.0 & 2.0 \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_posterior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Posterior predictive estimate for the core-S\'ersic fit for highest (fiducial) mass resolution. Darker shaded regions correspond to higher Bayesian Credible Intervals (BCIs). The observed data is overlaid, and seen to lie well within the 50\% BCI. }}
    \label{fig:postpred}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001_latentqty_compare}.pdf}
    \caption{\drafting{Latent params, and compare to naive estimates using Gaussian stats. Note error bars on black points are smaller than the point itself if not visible}}
    \label{fig:cs_compare_naive}
\end{figure*}

>>>>

\subsection{Projected Mass Density Profiles}
We use hierarchical modelling to recover the population parameter distributions of the core-S\'ersic model \autoref{eq:coresersic} for each simulation family. 
A probabilistic graphical model detailing the connection between hyperparameters, latent parameters, and observables is given in \autoref{fig:graphicalmodel}.

\begin{table}
    \centering
    \caption{The parameters for each hyperparameter distribution for the core-S\'ersic model, where each distribution is a gamma distribution.}
    \label{tab:cshypers}
    \begin{tabular}{ccc}
        \hline
        Hyperparameter & $\mu$ & $\sigma$ \\
        \hline
        $r_{\mathrm{b}, \mu}$ & 0.0 & 2.0 \\
        $r_{\mathrm{b}, \sigma}$ & 0.0 & 1.0 \\
        $R_{\mathrm{e}, \mu}$ & 0.0 & 14.0 \\
        $R_{\mathrm{e}, \sigma}$ & 0.0 & 6.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \mu}$ & 0.0 & 20.0 \\
        $\tilde{\Sigma}_{\mathrm{b}, \sigma}$ & 0.0 & 10.0 \\
        $\gamma_\mu$ & 0.0 & 0.5 \\
        $\gamma_\sigma$ & 0.0 & 1.0 \\
        $n_\mu$ & 0.0 & 8.0 \\
        $n_\sigma$ & 0.0 & 5.0 \\
        $a_\mu$ & 0.0 & 20.0 \\
        $a_\sigma$ & 0.0 & 10.0 \\
        \hline
    \end{tabular}
\end{table}

@@

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_posterior_pred_log10_proj_density_mean}.pdf}
    \caption{\drafting{Posterior predictive estimate for the core-S\'ersic fit for highest (fiducial) mass resolution. Darker shaded regions correspond to higher Bayesian Credible Intervals (BCIs). The observed data is overlaid, and seen to lie well within the 50\% BCI. }}
    \label{fig:postpred}
\end{figure*}

\begin{figure*}
    \includegraphics{{graham_density-H-H-3.0-0.001-hierarchy_latentqty_compare}.pdf}
    \caption{\drafting{Latent params, and compare to naive estimates using Gaussian stats. Note error bars on black points are smaller than the point itself if not visible}}
    \label{fig:cs_compare_naive}
\end{figure*}

### 2023-02-21 papers/paper-1/sections/stats.tex
Small updates in preparation for SPZ meeting

\subsection{Hamiltonian Monte Carlo}
An excellent overview of HMC may be found in \citet{betancourt17}, with a brief description provided here. \drafting{Mention NUTS here somewhere?}

In numerically sampling an arbitrary target distribution, we are required to sample from the \textit{typical set}, or those regions of the target distribution which provide the dominant contribution to an expectation value. 
Simplistically, this may be achieved using any Markov generating process, such as the widely-used Metropolis-Hastings (MH) algorithm, albeit at the expense of inefficient and biased sampling. 
The MH algorithm generates from a starting position $q$ a proposal for the next position $q'$, with proposals accepted with some probability $Q(q'|q)$, where $Q$ is often chosen to be the normal distribution. 
Sampling with increased efficiency can be achieved by using information about the geometry of the typical set of the target distribution through a vector field aligned with the typical set, which may be estimated using the gradient of the target distribution. 
To align the gradient vector field with the typical set vector field, we can introduce auxillary momenta variables $p$, promoting our $D$-dimensional target distribution to a projection of a $2D$-dimensional canonical distribution in some phase space:
\begin{equation}\label{eq:hmc}
    P(q,p) = P(p|q)P(q) = e^{-H(q,p)}
\end{equation}
where $H$ is an invariant Hamiltonian function. 
The desired vector field may then be generated from the Hamiltonian $H$ using Hamilton's equations. 
This formulation also lends itself to improved sampling efficiency by decomposing the Hamiltonian into a \textit{kinetic energy} term and a \textit{potential energy} term, with the latter equivalent to the logarithm of the target distribution. 
\begin{align}
    H(q,p) &= -\log P(q,p) \\
    &= -\log \left( P(p|q)P(q) \right) \\
    &= -\log P(p|q) -\log P(q)
\end{align}
As the Hamiltonian is phase space conserving, the Hamilton equations may be accurately integrated using a volume-preserving integration scheme, such as symplectic integrators. 
Of these integrators, the leapfrog scheme offers both a simple implementation and high accuracy over long integration times.

@@

To answer these questions, we define a test statistic $T$ from which we determine a Bayes-equivalent of the frequentist $p$-value \citep{gelman15}.
\begin{align}
    p_\mathrm{Bayes} &= \mathrm{Pr}\left( T(y^\mathrm{rep}, \theta) \geq T(y, \theta)|y \right) \\
    &= \iint I_{T(y^\mathrm{rep}, \theta) \geq T(y, \theta)|y} p\left( y^\mathrm{rep}|\theta\right) p\left( \theta|y\right) \dd{y^\mathrm{rep}} \dd{\theta}
\end{align}
Here, $I$ is the indicator function.

>>>>

@@

%To answer these questions, we define a test statistic $T$ from which we determine a Bayes-equivalent of the frequentist $p$-value \citep{gelman15}.
%\begin{align}
%    p_\mathrm{Bayes} &= \mathrm{Pr}\left( T(y^\mathrm{rep}, \theta) \geq T(y, \theta)|y \right) \\
%    &= \iint I_{T(y^\mathrm{rep}, \theta) \geq T(y, \theta)|y} p\left( y^\mathrm{rep}|\theta\right) p\left( \theta|y\right) \dd{y^\mathrm{rep}} \dd{\theta}
%\end{align}
%Here, $I$ is the indicator function.
